---
title: "Govindarajulu Distribution in Stan"
author: "Deemah"
date: "2/14/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
bibliography: "`r rbbt::bbt_write_bib('references.bib', overwrite = TRUE)`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Flexible quantile density function is
$$q_X(u)=Ku^\alpha(1-u)^\beta, \quad K>0;  \; \alpha, \beta \; \text{real}$$
This distribution has various names depending on values of parameters. We will
be looking at the distribution with
$\alpha=\gamma>0, \; \beta=1, \; K=\sigma\alpha(\alpha+1)$

This is Govindarajalu ditribution with quantile function

$$Q_X(u)=\sigma((\alpha+1)u^\alpha-\alpha u^{\alpha+1})$$

and quantile density function

$$q_X(u)=Ku^\alpha(1-u)=\sigma\alpha(\alpha-1)u^\alpha(1-u)$$ We will keep the
alpha-sigma parametrization, even though some literature uses alpha-K
parametrization.

In order to keep our functions versatile, let's make both vectorized and
non-vectorized versions. We will differentiate them by the suffix `_s` and `_v`
for scalar and vector, respectively.

Need to make:

-   `_cdf` equivalent, quantile function, to be named `govindarajulu_qf_s_cdf()`
    and `govindarajulu_qf_v_cdf()`
-   `_lcdf` equivalent, log quantile function, to be named
    `govindarajulu_qf_s_lcdf()` and `govindarajulu_qf_v_lcdf()`
-   `_lpdf` equivalent, log quantile density function, to be named
    `govindarajulu_qdf_s_lpdf()` and `govindarajulu_qdf_s_lpdf()` There's no
    equivalent to `_lccdf`?
-   `_rng` function is always scalar, which we will call
    `govindarajulu_q_s_rng()`
    
# Govindarajulu distribution

```{r, echo=FALSE}
library(rstan)
library(magrittr)
stan_code <- tempfile("govindarajulu", fileext = ".stan")
```

```{cat, echo=FALSE, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
functions{ // begin the functions block
```

## qf (cdf)

Govindarajulu quantile function from @nair2020BayesianInferenceQuantile

$$Q_X(u)=\sigma((\alpha+1)u^\alpha-\alpha u^{\alpha+1})=\sigma a u^a(1+1/a-u)$$

Here for vectorized function we could use the trick that `pow(a,b)` is
`exp(b*log(a))`. However we need to be aware that this can create errors for
invalid `log(a)`, so we will have to work around that. The vectorized version of
the power function has been cooking forever in Stan. Until it becomes available,
we will have to use loops

```{cat qfcdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
  real govindarajulu_qf_s_cdf(real p, real alpha, real sigma) {
    real inv_alpha = inv(alpha);
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
  return sigma*alpha*pow(p, alpha)*(1+inv_alpha-p);
  }

  real govindarajulu_qf_v_cdf(vector p, real alpha, real sigma) {
    int N = rows(p);
    real inv_alpha=inv(alpha);
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
   return exp(N*log(sigma)+N*log(alpha)+alpha*sum(log(p))+sum(log1p(inv_alpha-p)));
  }
  
  real govindarajulu_rng(real alpha, real sigma) {
    return metalog_qf_s_cdf(uniform_rng(0, 1), alpha, sigma);
}
```

## qf (lcdf)

Log of quantile function

$$log(Q_X(u))=ln(\sigma a u^a(1+1/a-u))=ln(\sigma)+ln(\alpha)+\alpha ln(u)+ln(1+1/a-u)$$

```{cat qflcdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
  real govindarajulu_qf_s_lcdf(real p, real alpha, real sigma) {
    real inv_alpha=inv(alpha);
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
  return log(sigma)+log(alpha)+alpha*log(p)+log(1+inv_alpha-p);
  }

  real govindarajulu_qf_v_lcdf(vector p, real alpha, real sigma) {
    int N = rows(p);
    real inv_alpha=inv(alpha);
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
   return N*log(sigma)+N*log(alpha)+alpha*sum(log(p))+sum(log1p(inv_alpha-p));
  }
```

## qdf (pdf)

Quantile density function

$$q_X(u)=\sigma\alpha(\alpha-1)u^\alpha(1-u)$$

```{cat qdfpdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
  real govindarajulu_qdf_s_pdf(real p, real alpha, real sigma) {
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
  return sigma*alpha*(alpha-1)+pow(p, alpha)*(1-p);
  }

  real govindarajulu_qdf_v_pdf(vector p, real alpha, real sigma) {
    int N = rows(p);
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
   return exp(N*log(sigma)+N*log(alpha)+N*log(alpha-1)+alpha*sum(log(p))+sum(log1m(p)));
  }
```

## qdf (lpdf)

Quantile density function

$$q_X(u)=\sigma\alpha(\alpha-1)u^\alpha(1-u)$$

Log quantile density function

$$ln(q_X(u))=ln(\sigma)+ln(\alpha)+ln(\alpha-1)+\alpha*ln(u)+ln(1-u)$$

```{cat qdflpdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
  real govindarajulu_qdf_s_lpdf(real p, real alpha, real sigma) {
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
  return log(sigma)+log(alpha)+log(alpha-1)+alpha*log(p)+log(1-p);
  }
  real govindarajulu_qdf_v_lpdf(vector p, real alpha, real sigma) {
    int N = rows(p);
    real inv_alpha=inv(alpha);
    vector[N] pow_p_alpha;
    // govindarajulu quantile function
    if (alpha<0)
      reject("alpha<0; found alpha =", alpha);
    if (sigma<0)
      reject("sigma<0; found sigma =", sigma);
   return N*log(sigma)+N*log(alpha)+N*log(alpha-1)+alpha*sum(log(p))+sum(log1m(p));
  }
```

# Generalized Exponential distribution

## cdf

Generalized exponential distribution CDF
$$F_Y(y|\alpha,\lambda)=(1-e^{-\lambda y})^{\alpha}$$
where $\alpha>0, \lambda>0$ are shape and scale parameters, respectively.

```{cat genexpcdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
//////// generalized exponential CDF ////////////
  real genexp_s_cdf(real x, real alpha, real lambda) {
    // generalized exponential cdf 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return (1-exp(-lambda*x))^alpha;
  }
  real genexp_v_cdf(vector x, real alpha, real lambda)  {
    int N = rows(x);
    real inv_alpha=inv(alpha);
    // govindarajulu quantile function
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
   return exp(alpha*sum(log1m(exp(-lambda*x))));
  }
```

## qf (cdf)

Generalized exponential distribution quantile function
$$Q_Y(v|\alpha,\lambda)=-\frac{1}{\lambda} ln(1-v^{1/\alpha})$$
where $\alpha>0, \lambda>0$ are shape and scale parameters, respectively.

```{cat genexpqf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
//////// generalized exponential QF function ////////////
 real genexp_qf_s_cdf(real p, real alpha, real lambda) {
    real inv_lambda = inv(lambda);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return -inv_lambda*log1m(pow(p,inv_alpha));
  }
  real genexp_qf_v_cdf(vector p, real alpha, real lambda)  {
    int N = rows(p);
    real inv_lambda = inv(lambda);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return -inv_lambda*log1m_exp(inv_alpha*sum(log(p)));
  }
//////// generalized exponential LOG QF function ////////////
  real genexp_qf_s_lcdf(real p, real alpha, real lambda) {
    real inv_lambda = inv(lambda);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return log(-inv_lambda)+log(log1m(pow(p,inv_alpha)));
  }
  real genexp_qf_v_lcdf(vector p, real alpha, real lambda)  {
    int N = rows(p);
    real inv_lambda = inv(lambda);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return log(-inv_lambda)*log(log1m_exp(inv_alpha*sum(log(p))));
  }
```

## qdf (pdf)

Generalized exponential distribution quantile function
$$q_Y(v|\alpha,\lambda)=\frac{d}{dv}(-\frac{1}{\lambda} ln(1-v^{1/\alpha}))=-1/\lambda*\frac{1}{(1-v^{1/\alpha})}*\frac{d}{dx}(1-x^{1/a})=$$
$$=\frac{d}{dx}[1]-\frac{d}{dx}[x^{1/\alpha}]*\frac{1}{\lambda(1-x^{1/a})}=\frac{0-\frac{1}{a}x^{1/a-1}}{\lambda(1-x^{1/a})}=\frac{x^{1/a-1}}{\alpha\lambda(1-x^{1/\alpha})}$$

where $\alpha>0, \lambda>0$ are shape and scale parameters, respectively.

```{cat genexpqdf, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
//////// generalized exponential QDF function /
  real genexp_qdf_s_pdf(real p, real alpha, real lambda) {
    real inv_alpha =inv(alpha);
    // generalized exponential quantile density function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return pow(p,inv_alpha-1)/(alpha*lambda*(1-pow(p, inv_alpha)));
  }
  real genexp_qdf_v_pdf(vector p, real alpha, real lambda)  {
    int N = rows(p);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile density function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return exp((inv_alpha-1)*sum(log(p))) - log(alpha)-log(lambda)-log1m(inv_alpha*sum(log(p)));
  }
//////// generalized exponential LOG QDF function ////////////
  real genexp_qdf_s_lpdf(real p, real alpha, real lambda) {
    real inv_alpha =inv(alpha);
    // generalized exponential quantile density function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return pow(p,inv_alpha-1)/(alpha*lambda*(1-pow(p, inv_alpha)));
  }
  real genexp_qdf_v_lpdf(vector p, real alpha, real lambda)  {
    int N = rows(p);
    real inv_alpha =inv(alpha);
    // generalized exponential quantile density function 
    if (alpha<=0)
      reject("alpha<=0; found alpha =", alpha);
    if (lambda<=0)
      reject("lambda<=0; found lambda =", lambda);
  return exp((inv_alpha-1)*sum(log(p))) - log(alpha)-log(lambda)-sum(log1m(inv_alpha*log(p)));
  }
```


```{cat, echo=FALSE, engine.opts=list(file=stan_code, append = TRUE), class.source="stan"}
} // finish the functions block
```

```{r printout, echo=FALSE}
writeLines(readLines(stan_code))
readr::write_lines(readLines(stan_code), "govi.stan")
```

Finish and test

```{r}
expose_stan_functions("govi.stan")

a <- 2.0915
sg <- 604.319/(a*(a+1))
govindarajulu_qdf_s_pdf(0.5, a, sg)

seq(0,1,by=0.001) %>% 
   sapply(govindarajulu_qf_s_cdf,a, sg) %>% 
  plot()
 seq(0,1,by=0.001) %>% 
   sapply(govindarajulu_qdf_s_pdf,a, sg) %>% 
   plot()

lam  <- 1

 seq(0,1,by=0.001) %>% 
   sapply(genexp_qf_s_cdf, a, lam) %>% 
   plot()

 genexp_qf_s_cdf(0.5, a, lam)
 
 seq(0,1,by=0.001) %>% 
   sapply(genexp_qdf_s_pdf, a, lam) %>% 
   plot()

govindarajulu_qf_s_cdf(0.5, a, sg)

```

```{stan output.var="govi_stan_model"}
#include govi.stan
data{
  int N;
  vector[N] u;
  real theta;
  real sigma;
}
parameters{
  real a;
}
model{
  target += genexp_qdf_s_lpdf(a | theta, 1);
  for(i in 1:N){
    target += govindarajulu_qdf_s_lpdf(u[i] | a, sigma);
  }
}
```

Prepare data and sample

```{r}
library(qpd)
vec_times <- c(0.1,7,36,67,84,0.2,11,40,67,84,1,12,45,67,84,1,18,46,67,85,1,18,47,72,85,1,18, 50,75,85,1,18,55,79,85,2,18,60,82,85,3,21,63,82,86,6,32,63,83,86)
vec_times_ord <- vec_times[order(vec_times)]

a=2.0915
sg=604.319/(a*(a+1))

afun <- function(u0, x_r, sg, a){
 tmp_u_r <- u0 + (x_r-qgovindarajulu(u0, sg, a))/fgovindarajulu(u0, sg, a)
 tmp_u_r[tmp_u_r<0] <- 0
 tmp_u_r[tmp_u_r>1] <- 1
 qgovindarajulu(tmp_u_r, sg, a)-x_r
}

u0s <- double(length = length(vec_times_ord))
for (r in seq_along(vec_times_ord)){
  tmp_u0 <- NULL
  tmp_u0 <- uniroot(afun, x_r=vec_times_ord[r], sg=sg, a=a, interval = c(0,1), tol = 1e-15)
  u0s[r] <- ifelse(is.null(tmp_u0), NA, tmp_u0$root)
}

dt_lst <- list(N=length(u0s), u=u0s, theta=a, sigma=1.265528)
fit <- sampling(govi_stan_model, data=dt_lst)
print(fit)
```




# References
