---
title: "Quantile Parametrized Bayesian Updating"
bibliography: "`r rbbt::bbt_write_bib('qpd-bayesian.bib', overwrite = TRUE)`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(qpd)
library(tidyverse)
library(hrbrthemes)
library(gt)
library(rmetalog)

library(rstan) 
rstan_options(auto_write = TRUE)
options(mc.cores = ceiling(parallel::detectCores()/2))
SEED <- 42 # set random seed for reproducability

fishSize <- rmetalog::fishSize
```

## Glossary

**Cumulative distribution function (CDF)** denoted by $F_X(x|\theta)$ is defined as probability of the random variable $X$ being less than or equal to some given value $x$.
$$F_X(x|\theta) = P(X \leq x|\theta)$$
**Probability density function (PDF)**, denoted by $f_X(x)$ is such that: 
$$f_X(x|\theta)dx=P(x\leq X \leq x+dx|\theta)=F_X(x+dx|\theta)-F_X(x|\theta)=\frac{dF_X(x|\theta)}{dx}$$
where $dx$ is infinitesimally small range of $x$. So PDF, $f_X(x)=dF_X(x)/dx$ is the first derivative of CDF and the area under the PDF curve $f_X(x)$ is equal to 1. 

**Quantile function (QF)**, denoted by $Q_X(u|\theta)$ is related to the probability $u$ defined as $u \text{ for which } P(X \leq u|\theta)=u$. The function $Q_X(u|\theta)$ is called a *quantile function* and it expresses *u-quantile* as a function of *u*. 

$$Q_X(u|\theta)=inf\{u:F_X(x|\theta)\geq u\}, \quad 0 \leq u \leq 1$$

For any pair of values along CDF $(x,p)$ can be written as $x=Q(p)$ and $p=F(x)$, therefore these functions are inverses of each other $Q(x)=F^{-1}(x)$ and $F(x)=Q^{-1}(x)$.

**Quantile density function (QDF)** denoted by $q_X(p)$ is the first derivative of the Quantile Function (QF), the same way as PDF is a first derivative of CDF.
$$q_X(p)=\frac{dQ_X(p)}{dp}$$
Because $q_X(p)$ is the slope of QF, it is non-negative on $0 \leq p \leq 1$.

The advantage of quantile functions is that "quantile functions can be added and, in the right circumstances, multiplied to obtain new quantile functions. Quantile density functions can also be added together to derive new quantile density functions" [@gilchrist2000StatisticalModellingQuantile].

## Important relationships

$$F(Q(u))=u$$
$$q(u)f(Q(u))=1$$
Hazard rate

$$h(x)=f(x)/\bar{F}(x))$$
where $\bar{F}(x)=1-F(x)$ is survival function of $x$ and $F(x)$ is a distribution function (CDF).

Also, hazard quantile function

  $$H(u)=h(Q(u))=[(1-u)q(u)]^{-1}$$
  To learn more about hazard quantile function see @nair2009QuantileBasedReliabilityAnalysis.

## Bayesian updating for quantile functions

The information about the parameter $\theta$ defined on $\theta\in A\subset\mathbb{R}$ can be described by prior distribution of random variable $\Theta$  with $F_\Theta(\theta)=v$.

Let's denote the sample of $x$ (of size $n$) from $F_X(x|\theta)$ as $\underline{x}=(x_1, x_2, ...,x_n)$. 

The posterior density $f_\Theta(\theta|\underline{x})$ is given by:

$$f_{\Theta|\underline{x}}(\theta|\underline{x})=K(\underline{x})f_X(\underline{x}|\theta)f_\Theta(\theta)$$
where $f_X(\underline{x}|\theta)=\prod_{i=1}^{n}f_X(x_i|\theta)$ is the likelihood and $K(\underline{x})$ is the normalizing constant, given by
$$(K(\underline{x}))^{-1}=\int_A f_X(\underline{x}|\theta)f_\Theta(\theta)d\theta$$

The posterior $f_{\Theta|\underline{x}}(\theta|\underline{x})$ can be expressed in terms of quantile function $\underline{Q}$ of the sample $\underline{x}$.

$$f_{\Theta|\underline{Q}}(\theta)=K(\underline{Q})f_X(\underline{Q}|\theta)f_\Theta(\theta)$$
where $\underline{Q}=(Q_1(u_1), Q_2(u_2), ...,Q_n(u_n))$ and $F(x_i|\theta)=u_i$ for $i=1,2,...,n$.

Because $F_\Theta(\theta)=v$ then $Q_\Theta(v)=\theta$. When CDF $F$ is continuous and strictly increasing (proper CDF) with quantile function $Q$, then $F(Q(u))=u$ and therefore 
$$f(Q(u))=\frac{dF(Q(u))}{dQ(u)} = \frac{dF(Q(u))/du}{dQ(u)/du}=\frac{dF(F^{-1}(u))/du}{q(u)}=\frac{du/du}{q(u)}=  (q(u))^{-1}$$ 
For the prior $f_{\Theta|\underline{Q}}(\theta)$ it would mean

$$f_\Theta(\theta)=\frac{dF_\Theta(\theta)}{d\theta}=\frac{dF_\Theta(Q_\Theta(v))}{d(Q_\Theta(v))}=\frac{dF_\Theta(Q_\Theta(v))/dv}{d(Q_\Theta(v))/dv}=\frac{dF_\Theta(F^{-1}_\Theta(v))/dv}{q_\Theta(v)}=\frac{dv/dv}{q_\Theta(v)}=\frac{1}{q_\Theta(v)}=(q_\Theta(v))^{-1}$$

We can substitute these into the formula for the posterior:

$$f_{\Theta|\underline{Q}}(Q_\Theta(v)|\underline{Q})=K_1(\underline{Q})f_X(\underline{Q}|Q_\Theta(v))(q_\Theta(v))^{-1}$$
where $0\leq u_i, \, v\leq 1$ and $(K_1(\underline{Q}))^{-1}=\int_{0}^{1}f_X(\underline{Q}|Q_\Theta(v))(q_\Theta(v))^{-1} dv$

Therefore posterior QDF of $\Theta|\underline{Q}$ has the form:

$$q_{\Theta|\underline{Q}}(v)=\frac{(K_1(\underline{Q}))^{-1}q_\Theta(v)}{f(\underline{Q}|Q_\Theta(v))}=(K_1(\underline{Q}))^{-1}q_\Theta(v)\prod_{i=1}^{n}q_X(u_i|Q_\Theta(v))$$
This last equation is the quantile version of the Bayes theorem which does not require the density function of the variable $X$.

Because
$$F_{\Theta|\underline{Q}}(Q_\Theta(v))=K_1(\underline{Q})\int_{0}^{v}f(\underline{Q}|Q_\Theta(p))dp$$
The right hand side of this equation can be denoted as $H(v)$ and it is a distribution on $[0,1]$ and therefore

$$Q_\Theta(v)=Q_{\Theta|\underline{Q}}(H(v))$$
Thus the posterior can be expressed as

$$Q_{\Theta|\underline{Q}}(v)=Q_\Theta(H^{-1}(v))$$
The final equation expresses the relationship between the prior and the posterior quantile function and how the prior distribution of $\theta$ changes in light of observations on $X$.

@nair2020BayesianInferenceQuantile also use alternative expression for H(v):

$$H(v)=K_2\int_{0}^{u}(q(\underline{Q}|Q_{\theta}(p)))^{-1}dp$$
where $K_2^{-1}=\int_0^{1}(q(\underline{Q}|Q_{\theta}(p)))^{-1}dp$

Equations [ref] above are expressed in terms of $u_i=F(x_i|\theta)$. In reality only observational values of $\underline{x}$ are known. They can be ordered so that the $r$-th order statistic $x_{(r)}=Q_X(u_{(r)|\theta})$ for every $r = 1,2,...n$. In order to transition from $x_{(r)}$ to $u_{(r)}$, we need to have some value of $0<u_0<1$ 

$$Q_X(u_{(r)})=x_{(r)}=Q_X(u_0)+(u_{(r)}-u_0)q_X(u_0)$$ 
giving

$$u_{(r)}=\frac{x_{(r)}-Q_X(u_0)}{q_X(u_0)}+u_0$$
 The initial $Q_X(u_0)$ is obtained by replacing $\theta$ in it by the suitable single point-estimate derived from prior distribution of $\theta$ (mean or median). Once $u_{(r)}$ is obtained from the equation above, it replaces $u_0$ and the process is repeated by successively changing $u_{(r)}$ in previous iteration until $|Q(u_{(r)})-x_{(r)}|<\epsilon$, a predefined small quantity.

# Running QPD distributions in Stan

In order to run the functions provided in this package in stan, the one must specify the following in the chunk options:
`{stan, output.var="out_var", engine.opts=list(isystem=system.file("stan", package="qpd"))}` and include `#include qpd.stan` in the header of the Stan chunk.

Let's try and implement the simplest quantile function in Stan. The set of functions that need to implemented includes:

- `_lpdf`,`_lcdf`, `_lccdf` as sum of log values for each element of Y
- `_cdf` as product of values for each element of Y
- `_rng` single number drawn from distribution

# Example 1

Data from @lai2006StochasticAgeingDependence. 

```{r}
library(magrittr)

vec_times <- c(0.1,7,36,67,84,0.2,11,40,67,84,1,12,45,67,84,1,18,46,67,85,1,18,47,72,85,1,18, 50,75,85,1,18,55,79,85,2,18,60,82,85,3,21,63,82,86,6,32,63,83,86)
#vec_times2 <- c(0.9, 1.5, 2.3, 3.2, 3.9, 5.0, 6.2, 7.5, 8.3, 10.4, 11.1, 12.6, 15.0, 16.3, 19.3, 22.6, 24.8, 31.5, 38.1, 53.0)

vec_times_ord <- vec_times[order(vec_times)]
#vec_times_ord <- vec_times2[order(vec_times2)]
```

## Model

Flexible quantile density function is 
$$q_X(u)=Ku^\alpha(1-u)^\beta, \quad K>0; \alpha, \beta \; \text{real}$$
This distribution has various names depending on values of parameters. We will be looking at the distribution with $\alpha=\gamma>0, \; \beta=1, \; K=\sigma\alpha(\alpha+1)$

This is Govindarajalu ditribution with quantile function 

$$Q_X(u)=\sigma((\alpha+1)u^\alpha-\alpha u^{\alpha+1})$$ 

The authors discuss estimating of $\alpha$ and $\K$ using L-moments with $\alpha^{*}=2.0915$ and $K^{*}=604.319$.

```{r}
a=2.0915
sg=604.319/(a*(a+1))

#u0 <- seq(0,1,by=0.01) #u0 <- 0.02
afun <- function(u0, x_r, sg, a){
 tmp_u_r <- u0 + (x_r-qgovindarajulu(u0, sg, a))/fgovindarajulu(u0, sg, a)
 tmp_u_r[tmp_u_r<0] <- 0
 tmp_u_r[tmp_u_r>1] <- 1
 qgovindarajulu(tmp_u_r, sg, a)-x_r
}

u0s <- double(length = length(vec_times_ord))
for (r in seq_along(vec_times_ord)){
  tmp_u0 <- NULL
  tmp_u0 <- uniroot(afun, x_r=vec_times_ord[r], sg=sg, a=a, interval = c(0,1), tol = 1e-15)
  u0s[r] <- ifelse(is.null(tmp_u0), NA, tmp_u0$root)
}
```


```{r, eval=FALSE}
library(rstan)
library(shinystan)
#govi_model <- system.file("stan/govi.stan", package="qpd")
govi_model <- "../inst/stan/govi.stan"

gov <- stan_model(govi_model)

theta_est=log(1-exp(-a))/log(0.5)

dt_lst <- list(y=u0s, N=length(u0s), sg=sg, theta=theta_est)

st_res <- sampling(gov, data=dt_lst, iter=5000)
print(st_res)

shinystan::launch_shinystan(st_res)

# transformed data (into quantiles)
#u0s

# non-informative prior distribution of f(a) = 1/C (0<a<C) then prior quantile function is 
#Qa(v) =C*v, 0<v<1




art_res_1 <-c(0.022306,0.068111,0.237570,0.320954,0.515055,
0.659505,0.759336,0.809422,0.832348,0.0311644,0.0957779,
0.295004,0.407408,0.536175,0.659505,0.788389,0.820606,0.832348,
0.0681112,0.117136,0.295004,0.43678,0.571534,0.659505,
0.788389,0.820606,0.0681116,0.166043,0.295004,0.465539,
0.607435,0.659505,0.798706,0.820606,0.0681116,0.179636,
0.295004,0.500966,0.629422,0.69887,0.809422,0.820606,
0.0681116,0.226963,0.295004,0.508014,0.629422,0.723818,0.809422,
0.820606)
sort(art_res_1)
prod(art_res_1)^-1


plot(u0s-sort(art_res_1))
abline(a=0, b=0)

```


# References
