---
title: "Stan exp signature"
author: "Deemah"
date: "2/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
```

# Original

```{stan, output.var="orig_stan"}
data{
  int LENGTH;
  vector [LENGTH] Y;
}
parameters{
  real<lower=0> lambda;
}
model{
  lambda ~ gamma(1.0, 1.0);
  Y ~ exponential(lambda);
}
generated quantities{
  real pred;
  pred = exponential_rng(lambda);
}
```

Test it

```{r}
lambda <- 1
dat <- rexp(500, lambda)
len <- length(dat)
              
fit <- sampling(orig_stan, data=list(Y=dat, LENGTH=len))
print(fit)
```

# Custom likelihood

I am going to define custom functions both with the first argument of type `vector`. We will then replace the buil-in exponential likelihood with our own custom log likelihood function. 

```{stan, output.var="custlik_stan"}
functions{
  real dv_exponential_lpdf(vector x, real lam){
    return sum(log(lam)-x*lam);
  }
  real dv_gamma_lpdf(vector y, real alpha, real beta){
    real loglik;
    loglik = sum(alpha*log(beta)+(alpha-1)*log(y)-beta*y-tgamma(alpha));
    return loglik;
  }
}
data{
  int LENGTH;
  vector [LENGTH] Y;
}
parameters{
  real<lower=0> lambda;
}
model{
  lambda ~ gamma(1.0, 1.0);
  Y ~ dv_exponential(lambda);
}
generated quantities{
  real pred;
  pred = exponential_rng(lambda);
}
```

Test it

```{r}
lambda <- 1
dat <- rexp(500, lambda)
len <- length(dat)
              
fit <- sampling(custlik_stan, data=list(Y=dat, LENGTH=len))
print(fit)
```

However, if we try to replace the built-in `gamma()` with our own `dv_gamma()` we are going to get the following error message:

```{r}
#> No matches for: 
#> 
#>   real ~ dp_gamma(real, real)
#> 
#> Available argument signatures for dp_gamma:
#> 
#>   vector ~ dp_gamma(real, real)
```

This means that we can not use vectorized functions in place of the prior, and we have to rewrite our functions to accept only scalars.


```{stan, output.var="s_custlik_stan"}
functions{
  real ds_exponential_lpdf(real x, real lam){
    return log(lam)-x*lam;
  }
  real ds_gamma_lpdf(real y, real alpha, real beta){
    real loglik;
    loglik = alpha*log(beta)+(alpha-1)*log(y)-beta*y-tgamma(alpha);
    return loglik;
  }
}
data{
  int LENGTH;
  vector [LENGTH] Y;
}
parameters{
  real<lower=0> lambda;
}
model{
  target += ds_gamma_lpdf(lambda | 1.0, 1.0);

  for (i in 1:LENGTH){
    target += ds_exponential_lpdf(Y[i] | lambda);
 }
}
generated quantities{
  real pred;
  pred = exponential_rng(lambda);
}
```


```{r}
fit <- sampling(s_custlik_stan, data=list(Y=dat, LENGTH=len))
print(fit)
```

